---
output: github_document
date: "`r format(Sys.time(), '%B %d %Y')`"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

**IMPORTANT UPDATE: An updated version of this package will be released soon!**


gibbonR: An R package for classification, detection and visualization of
acoustic signals using machine learning
================
Dena J. Clink & Holger Klinck 

### Package description

The field of bioacoustics is inherently multidisciplinary and relies on
computer scientists, engineers, and ecologists. This package is directed
towards ecologists who are interested in incorporating bioacoustics into
their research, but may not have the skills or training. The goal for
the creation of this package was to make commonly used signal processing
techniques and various machine learning algorithms readily available in R for
anyone interested in using bioacoustics in their research.

```{r echo=FALSE,warning=FALSE,message=FALSE,error=FALSE, fig.keep='all'}
library(ggpubr)
mainImage <- magick::image_read('man/gibbonR_flowchart.png')
(mainImage)
```

### How to cite

This package is currently in development, with submission to rOpenSci planned shortly. In the interim, please cite the arXiv preprint:

Clink, D. J. & H. Klinck. (2019). gibbonR: An R package for the detection and classification of acoustic signals using machine learning. arXiv, 1906.02572.
https://doi.org/10.48550/arXiv.1906.02572
