---
title: "gibbonR: An R package for the automated detection and classification of female gibbon calls from long-term acoustic recordings"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Getting started
## You can install the development version from [GitHub](https://github.com/DenaJGibbon) with:
```{r, echo=T,warning=FALSE, results='hide'}
# install.packages("devtools")
# devtools::install_github("DenaJGibbon/gibbonR")

library(gibbonR)
```

# Part 1. Prepare Training Data
In 'gibbonR' there are two ways that you can format your training data. The first can be a set of labelled .wav clips with the class indicated in the name of the file (e.g., 'gibbon_01.wav' and 'noise_01.wav'). The second is to have a folder of selection tables created in Raven Pro (K. Lisa Yang Center for Conservation Bioacoustics) and a folder with the associated '.wav' files. For the second approach there must be an annotation column indicating the call type and it is assumed that all signals of interest are annotated, and the rest of the files contain only background noise.  

## Part 1A. Training Data with Labeled .wav clips
### Read in clips and calculate MFCCs
```{r echo = T, results = 'hide' }
TrainingWavFilesDir <- 
  "/Users/denaclink/Desktop/RStudio Projects/gibbonR/data/MultipleSoundClasses/"

trainingdata <- gibbonR::MFCCFunction(input.dir=TrainingWavFilesDir, min.freq = 400, max.freq = 1600,win.avg="TRUE")


trainingdata$class <- as.factor(trainingdata$class)
```

### Compare Random Forest and Support Vector Machine for Supervised Classification
```{r }

trainingdata$class <- as.factor(trainingdata$class)


ml.model.svm <- e1071::svm(trainingdata[, 2:ncol(trainingdata)], trainingdata$class, kernel = "radial", 
                           cross = 25,
                           probability = TRUE)

print(paste('SVM accuracy',ml.model.svm$tot.accuracy))


ml.model.rf <- randomForest::randomForest(x=trainingdata[, 2:ncol(trainingdata)], y = trainingdata$class)


print(ml.model.rf)
```

## Part 1B. Training Data with Raven Selection Tables
### Prepare training data from labeled annotations
```{r,eval=F }
# Specify the folder where the training data will be saved
TrainingDataFolderLocation <- "/Users/denaclink/Desktop/RStudio Projects/gibbonR/data/TrainingDataFromRavenSelectionTables"

# Directory with annotated selection tables
AnnotatedSelectionTables <- list.files("/Users/denaclink/Desktop/RStudio Projects/gibbonR/data/SelectionTables/GibbonTrainingSelectionTables/",
                                       full.names = T)

# Directory with corresponding .wav files
AnnotatedWaveFiles <- list.files("/Users/denaclink/Library/CloudStorage/Box-Box/gibbonRSampleFiles/GibbonTrainingFiles/",full.names = T)
AnnotatedWaveFilesShort <- list.files("/Users/denaclink/Library/CloudStorage/Box-Box/gibbonRSampleFiles/GibbonTrainingFiles/",full.names = F)
AnnotatedWaveFilesShort <- str_split_fixed(AnnotatedWaveFilesShort,pattern = '.wav', n=2)[,1]

# Loop to cut out the corresponding annotations into short clips
for(i in 1: length(AnnotatedSelectionTables)){
  
  # Read in selection table
  TempSelectionTable <- read.delim2(AnnotatedSelectionTables[i])
  
  # Find the corresponding soundfile
  SoundFileIndex <- which(str_detect(AnnotatedSelectionTables[i],AnnotatedWaveFilesShort))
  
  TempAnnotateWave <- readWave(AnnotatedWaveFiles[SoundFileIndex])
  
  ShortSoundClips <- lapply(1:nrow(TempSelectionTable),
                                function(j) extractWave(TempAnnotateWave,
                                                        from= as.numeric(TempSelectionTable[j,]$Begin.Time..s.),
                                                        to=as.numeric(TempSelectionTable[j,]$ End.Time..s.),
                                                        xunit = c("time"),plot=F,output="Wave"))
  # Write wave files to folder
  for(k in 1:length(ShortSoundClips)){
    TempClip <- ShortSoundClips[[k]]
    WavFileName <- paste(TrainingDataFolderLocation,'/female.gibbon_', k, '.wav',sep="")
    writeWave(TempClip,WavFileName,extensible = F)
  }
  
  
}

```

### Prepare noise training data from files without target signal
```{r eval=F}
# Specify the folder where the training data will be saved
TrainingDataFolderLocation <- "/Users/denaclink/Desktop/RStudio Projects/gibbonR/data/TrainingDataFromRavenSelectionTables/"

# Directory with annotated selection tables
NoiseSelectionTables <- list.files("/Users/denaclink/Desktop/RStudio Projects/gibbonR/data/SelectionTables/NoiseSelectionTables/",
                                       full.names = T)

# Directory with corresponding .wav files
NoiseWaveFiles <- list.files("/Users/denaclink/Library/CloudStorage/Box-Box/gibbonRSampleFiles/NoiseFiles/",full.names = T)
NoiseWaveFilesShort <- list.files("/Users/denaclink/Library/CloudStorage/Box-Box/gibbonRSampleFiles/NoiseFiles/",full.names = F)
NoiseWaveFilesShort <- str_split_fixed(NoiseWaveFilesShort,pattern = '.wav', n=2)[,1]

for(i in 1:length(NoiseSelectionTables)){
  
  # Find the corresponding soundfile
  SoundFileIndex <- which(str_detect(NoiseSelectionTables[i],NoiseWaveFilesShort))

  DetectBLED(input=NoiseWaveFiles[SoundFileIndex],
           min.freq = 400, 
           max.freq = 1600,
           noise.quantile.val=0.3,
           file.type='wav',
           spectrogram.window =512,
           pattern.split = ".wav", 
           min.signal.dur = 3,
           max.sound.event.dur = 12, 
           output = "wav",
           wav.output = "TRUE", 
           output.dir = TrainingDataFolderLocation,
           swift.time=TRUE,
           time.start=06,
           time.stop=11,
           write.csv.output=TRUE,
           verbose=TRUE,
           random.sample=FALSE)
}

```

### Now read in clips based on Raven Selection tables and calculate MFCCs
```{r echo = T, results = 'hide' }

TrainingWavFilesDir <- 
  "/Users/denaclink/Desktop/RStudio Projects/gibbonR/data/TrainingDataFromRavenSelectionTables/"

trainingdata <- gibbonR::MFCCFunction(input.dir=TrainingWavFilesDir, min.freq = 400, max.freq = 1600,win.avg="TRUE")


trainingdata$class <- as.factor(trainingdata$class)
```

### Compare Random Forest and Support Vector Machine for Supervised Classification
```{r }

trainingdata$class <- as.factor(trainingdata$class)


ml.model.svm <- e1071::svm(trainingdata[, 2:ncol(trainingdata)], trainingdata$class, kernel = "radial", 
                           cross = 25,
                           probability = TRUE)

print(paste('SVM accuracy',ml.model.svm$tot.accuracy))


ml.model.rf <- randomForest::randomForest(x=trainingdata[, 2:ncol(trainingdata)], y = trainingdata$class)


print(ml.model.rf)


```

# Part 2. Run the detector/classifier

## Part 2a. Feature extraction
```{r }
# Specify the folder where the training data will be saved
TrainingDataFolderLocation <- "/Users/denaclink/Desktop/RStudio Projects/gibbonR/data/TrainingDataFromRavenSelectionTables/"
  
TrainingDataMFCC <- MFCCFunction(input.dir= TrainingDataFolderLocation, min.freq = 400, max.freq = 1600,win.avg="TRUE")
  
  TrainingDataMFCC$class <- as.factor(TrainingDataMFCC$class)
```
  
## Part 2b. Run DetectClassify  
```{r }

  TestFileDirectory <- '/Users/denaclink/Library/CloudStorage/Box-Box/gibbonRSampleFiles/GibbonTestFiles'
  
  OutputDirectory <-  "/Users/denaclink/Desktop/RStudio Projects/gibbonR/data/DetectAndClassifyOutput"
  
  DetectAndClassify(input=TestFileDirectory,
                    input.type='directory',
                    feature.df=TrainingDataMFCC,
                    model.type.list=c('SVM','RF'),
                    tune = TRUE,
                    short.wav.duration=300,
                    target.signal = c("female.gibbon"),
                    min.freq = 400, max.freq = 1600,
                    noise.quantile.val=0.15,
                    time.window.number =3,
                    n.windows = 9, num.cep = 12,
                    spectrogram.window =160,
                    pattern.split = ".wav",
                    min.signal.dur = 3,
                    max.sound.event.dur = 25,
                    maximum.separation =1,
                    probability.thresh.svm = 0.15,
                    probability.thresh.rf = 0.15,
                    wav.output = "TRUE",
                    output.dir =OutputDirectory,
                    swift.time=TRUE,time.start=5,time.stop=10,
                    write.csv.output=FALSE,verbose=TRUE,
                    random.sample='NA')
  

```

# Part 3. Calculate performance metrics

## Part 3a. Prepare data for performance metrics
```{r }
# Set location of test file selection tables
input.dir.text.files <- "/Users/denaclink/Desktop/RStudio Projects/gibbonR/data/SelectionTables/GibbonTestSelectionTables"

Annotatedfiles <- list.files(input.dir.text.files,full.names = T)

ListOfAnnotatedFilesShort <- list.files(input.dir.text.files,full.names = F)

nslash <- str_count(Annotatedfiles,pattern = '/')[1]+1
snames <- str_split_fixed(Annotatedfiles,pattern = '/',n=nslash)[,nslash]

all.detections <- data.frame()
for(x in 1:length(Annotatedfiles)){
  temp.table <- read.delim2(Annotatedfiles[x],fill = T,header =T)
  file.name <- str_split_fixed(snames[x],pattern = '[.]',n=2)[,1]
  recorder <- str_split_fixed(file.name,pattern='_',n=3)[,1]
  date <- str_split_fixed(file.name,pattern='_',n=3)[,2]
  time <- str_split_fixed(file.name,pattern='_',n=3)[,3]
  
  if(nrow(temp.table >0)){
    temp.table.updated <- cbind.data.frame(file.name,recorder,date,time,temp.table)
  } else {
    temp.row <- as.data.frame(t(rep('NA',ncol(temp.table))))
    colnames(temp.row) <- colnames(temp.table)
    temp.table.updated <- cbind.data.frame(file.name,recorder,date,time,temp.row)
    
  }
  all.detections <- rbind.data.frame(all.detections,temp.table.updated)
}

```

## Part 3b. Identify true and false positives
```{r }
  OutputDirectory <-  "/Users/denaclink/Desktop/RStudio Projects/gibbonR/data/DetectAndClassifyOutput"
    
  all.combinedprecision.recall.randomiter <- data.frame()
  range.secs.start <- 6
  range.secs.end <- 6
  
  ### Detections using band-limited energy summation
  gibbondetects <- OutputDirectory
  list.ml <-  list.files(gibbondetects, full.names = T, pattern='.wav')

  
  # Need to focus on gibbons for this validation
  nslash <- str_count(list.ml[[1]],'/')+1
  list.ml.signals <- str_split_fixed(list.ml,pattern = '/',n=nslash)[,nslash]
  
  list.ml.signals <- str_split_fixed(list.ml.signals,pattern = '_',n=5)[,4]
  
  
  list.ml <- 
    list.ml[which(list.ml.signals=='female.gibbon')]
  
  
  ml.detection.df <- data.frame()
  
  for(y in 1:length(list.ml)){
    L.wav <- list.ml[[y]]
    n.slash  <- str_count(L.wav, pattern = "/")[1] + 1
    
    det.file.name <- str_split_fixed(L.wav,"/",n=n.slash)[,n.slash]
    det.file.name <- str_split_fixed(det.file.name,".wav",n=2)[,1]
    
    file.name <- paste(str_split_fixed(det.file.name,"_",n=5)[,1],str_split_fixed(det.file.name,"_",n=5)[,2],
                       str_split_fixed(det.file.name,"_",n=5)[,3], sep='_')
    det.date <- str_split_fixed(det.file.name,"_",n=5)[,2]
    det.time <- str_split_fixed(det.file.name,"_",n=5)[,3]
    det.swift <- str_split_fixed(det.file.name,"_",n=5)[,1]
    det.time.start <- as.numeric(str_split_fixed(det.file.name,"_",n=9)[,6])
    det.time.end <- as.numeric(str_split_fixed(det.file.name,"_",n=9)[,7])
    probability <- str_split_fixed(det.file.name,"_",n=8)[,8]
    ml.algorithm <- str_split_fixed(det.file.name,"_",n=7)[,5]
    
    detections.df <- cbind.data.frame(file.name,det.swift, det.date, det.time,det.time.start,det.time.end,probability,ml.algorithm)
    
    ml.detection.df <- rbind.data.frame(ml.detection.df,detections.df)
  }
  
  
  recall.snr.all.df <- data.frame()
  for(x in 1:nrow(ml.detection.df)){
    all.detections.subset <- ml.detection.df[x,]
    validate.detect.subset <-subset(all.detections,file.name==as.character(all.detections.subset$file.name))
    validate.detect.subset$Begin.Time..s. <- as.numeric(validate.detect.subset$Begin.Time..s.)
    min.start.time <- as.numeric(all.detections.subset$det.time.start)-range.secs.start
    max.start.time <- as.numeric(all.detections.subset$det.time.start)+range.secs.end
    
    detections.ml <- subset(validate.detect.subset, Begin.Time..s.>min.start.time & Begin.Time..s.< max.start.time)
    
      if(nrow(detections.ml)>0){
      all.detections.subset$class.label <- '1'
      } else{
        all.detections.subset$class.label <- '-1'
      }
   
    recall.snr.all.df <- rbind.data.frame(recall.snr.all.df,all.detections.subset)
  }
  

```
## Part 3c. Calculate and plot performance metrics using 'ROCR'

```{r }
library(ROCR)

auc.df <- data.frame()
performance.df <- data.frame()

  
  ml.index <- unique(recall.snr.all.df$ml.algorithm)
  for(m in 1:length(ml.index)){
  
    temp.subset <-
    subset(recall.snr.all.df,
           ml.algorithm==ml.index[m])
  
  predictions <- as.numeric(temp.subset$probability)
  labels <- (temp.subset$class.label)
  pred <- prediction(predictions, labels)
  perf <- performance(pred, "rec", "prec")
  perfauc <- performance(pred, "aucpr")
  Precision <- perf@x.values[[1]]
  Recall <- perf@y.values[[1]]
  Threshold <- perf@alpha.values[[1]]
  AUC <- perfauc@y.values[[1]]
  perfF1 <- performance(pred, "f")
  F1 <-  perfF1@y.values[[1]]
  print(AUC)
  ml.algorithm <- ml.index[m]
  tempauc <- cbind.data.frame(AUC,ml.algorithm)
  auc.df <- rbind.data.frame(auc.df,tempauc)
  
  temp.performance <- cbind.data.frame(Precision,Recall,Threshold,F1,ml.algorithm)
  performance.df <- rbind.data.frame(performance.df,temp.performance)
  
  perf <- performance(pred, "prec", "rec")

  plot(perf,
     avg= "threshold",
     colorize=TRUE,
     lwd= 3,
     main= paste(ml.index[m],'Precision/Recall'))
  
plot(perf,
     lty=3,
     col="grey78",
     add=TRUE)

  }  

```

# Part 4. Unsupervised clustering 
## Part 4a. Create a UMAP plot colored by class
```{r }
library(gibbonR)
library(ggpubr)
UMAPBiplotAddSpectrograms(input.dir.Focal="/Users/denaclink/Desktop/RStudio Projects/gibbonR/data/MultipleSoundClasses/",output.dir.Focal="/Users/denaclink/Desktop/RStudio Projects/gibbonR/data/MultipleSoundClasses/Thumbnails/",add.spectrograms=TRUE,min.freq=400,max.freq=1600,main="UMAP Plot")
  
```
## Part 4b. Create a UMAP plot colored by affinity propagation clustering
```{r }
library(gibbonR)
library(ggpubr)
library(apcluster)
AffinityBiplotAddSpectrograms(input.dir.Focal="/Users/denaclink/Desktop/RStudio Projects/gibbonR/data/MultipleSoundClasses/",output.dir.Focal="/Users/denaclink/Desktop/RStudio Projects/gibbonR/data/MultipleSoundClasses/Thumbnails/",class='fixed', q.fixed=0.1,add.spectrograms=TRUE,min.freq=400,max.freq=1600,main="UMAP Plot")
  
```

